Panarchy: ripples of a boundary concept

### 2021-09-15 paper/paper_panarchy.Rmd
Second round of review, updated analysis with lemmatization.

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry Peterson^1^
    | \small
    | \footnotesize ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \footnotesize ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \footnotesize ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \footnotesize ^4^Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \footnotesize ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \footnotesize ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, 
    | \footnotesize Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \footnotesize ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA
abstract: |
  How do social-ecological systems change over time? In 2002 Holling and colleagues proposed the concept of Panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines, and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by combining qualitative methods and machine learning. Document analysis was used to code panarchy features that are commonly used in the scientific literature (N = 42), a qualitative analysis that was complemented with topic modeling of 2177 documents. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.
bibliography: panarchy.bib
csl: "/Users/juanrocha/Documents/styles/ecology.csl"
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
---

@@

Panarchy proposes that it is useful to conceptualize systems in terms of
interacting adaptive cycles. The adaptive cycle was an idea first proposed by
Holling based upon his experience working and studying managed ecosystems
[@biosphere:hJ4c_Ua8]. It was meant to be a conceptual tool that focused
attention on processes of destruction and reorganization, which had been
neglected in comparison to those of growth and conservation. Considering these
processes provides a more complex way to understand system dynamics.

An adaptive cycle alternates between long periods of system aggregation,
connection, and accumulation, and shorter periods of disruption and
reorganization. The adaptive cycle exhibits four phases arranged into two major
loops (Fig. \ref{fig:fig1}). The phases are exploitation ($r$), conservation
($K$), release ($\Omega$), and reorganization ($\alpha$), the latter
characterized by events of creative destruction (Fig. \ref{fig:fig1}). The first
loop, often referred to as the front-loop, from $r$ to $K$, is the slow,
incremental phase of growth and accumulation. The second loop, referred to as
the back-loop, from $\Omega$ to $\alpha$, is the rapid phase of reorganization
leading to renewal. A system going into the back-loop can either remain in a
similar form, or may transition to a new type of system with new boundaries and
key components [@Holling:2002wu]. Resilience researchers proposed that tensions
between demands for organization and efficiency versus demands for novelty and
diversity drive adaptive cycle type dynamics in many different types of complex
systems [@Holling:2002wu].

Panarchy is not a theory of what it is, but a metaphor of what might be
[@Gunderson:2002vk]. It is not a predictive tool, but aims to understand
adaptive change. The adaptive cycle is hypothesized to exist within a
three-dimensional space defined by three properties: potential, connectivity,
and resilience [@Gunderson:2002vk]. Potential refers to the capital available to
the system, for example nutrients and carbon captured by a forest, or human
capital - skills and knowledge accumulated to run the economy. Connectivity is a
proxy of the structure of the system; it is the network of interactions and
strength between its elements. Resilience is the capacity of any system to
absorb disturbance and reorganize while undergoing change so as to retain
essentially the same identity, as determined by its function, structure, and
feedbacks [@Anonymous:2016fv].

@@

The three-dimensional space has corners with attractors that can cause the
adaptive cycle to stagnate: poverty and rigidity traps [@Gunderson:2002vk].
Poverty traps are described in panarchy as maladaptive states where potential,
connectivity and resilience are low. Poverty traps are a series of feedback
mechanisms that reinforce impoverished states [@Bowles:2006hx; @Maru:2012dm;
@Allison:2004te], limiting the system's capacity to innovate and increase
potential. The opposite corner, where potential, connectivity and resilience are
high, is another maladaptive space called a rigidity trap. In that corner there
is little space for experimentation and innovation. Examples include systems
where ecological resilience has been extensively replaced by artificial
processes to maintain the system such as levees, flood barriers, or chemical
control of pests [@Holling:1996wx].

Panarchy offers a rich conceptual framework for understanding environmental
problems. While inspired by several mathematical constructs (e.g. cycles, traps,
scaling laws), it is general enough to invite scholars from multiple
disciplinary backgrounds, ontologies and epistemologies, to collaborate around
research questions and applied problems. As such, it is useful as a boundary
object and can be used empirically or metaphorically. The book itself presents a
series of case studies where geographers, economists, political scientists, and
ecologists have demonstrated the utility of the framework to their area of
research. However, the extent to which these concepts have gained empirical
support beyond the metaphor, or how they have evolved over time, remain open
questions. For example, the adaptive cycle is a useful metaphor to look back at
history and organize events and periods around phases, but it remains
challenging to identify the phase of a system in real time, or to use panarchy
to project future trajectories.

@@

To answer these questions we combined an automated literature review based on
topic modeling [@Blei:2012dk; @Griffiths:2004ey] with human-coded document
analysis [@Bryman:2008va].

*Data*: We used the Web of Science, Scopus, and GoogleScholar to survey academic
literature that has used or referenced works that trace back to Panarchy
[@Gunderson:2002vk]. We extracted complete records from the Scopus database that
matched the search for *"panarchy OR adaptive cycle"* (N = 595), or the search
*"(panarchy OR adaptive cycle) AND resilience"* (N = 278). The data was combined
with all papers (N = 1923) that cited the inaugural paper that introduced the
book to the academic community [@Holling:2001cu]. Records with missing abstracts
were dropped (N = 191), and records with missing year were set to 2020 given
that they are accepted manuscripts with digital object identifier (doi)
scheduled to be published later in 2020 or 2021.

To prepare the data for topic modeling, we constructed a document term matrix
with documents in rows and words in columns. Here our unit of analysis for the
document are the abstracts retrieved, and the matrix contains the count of the number of occurrences of each word per abstract. We removed stop words (e.g. "the", "a") and digits from the
matrix, as well as a list of words that were over represented in our data and
are common in the scientific literature but are unrelated to the papers' topics
(e.g. "paper", "study", "aim").

@@

The underlying statistical technique for this machine learning approach is
called Latent Dirichlet Allocation (LDA) [@Blei:2003tn]. It allocates
probabilities to latent variables (topics) based on the distribution of words in
text data, assuming a multivariate continuous (Dirichlet) distribution. We
compared three LDA algorithms: correlated topic models (CTM), variational
expectation maximization (VEM), and Gibbs sampling (Gibbs), by assessing their
performance against their log-likelihood estimation, entropy and perplexity
[@Grun:2011tb]. Entropy is a measure of order or disorder of a system; in the
context of topic models it measures how evenly the topic distribution is spread.
Perplexity measures the uncertainty of predicting a single word, so if the model
performance were the same as random, perplexity would approximate the vocabulary
size (N = 16 728 words). These performance metrics enable us to choose which
algorithm best fits our data, what is the optimal number of topics to fit, and
how to avoid overfitting.

*Document analysis*: We complemented our unsupervised approach with document
analysis [@Bryman:2008va] by coding an additional set of categorical variables
for a sample of papers (N = 42). Based on the reading of Panarchy, we first
developed a series of categorical variables that captured the main components of
the scientific framework. The variables were presented and received feedback at
the science meeting of the Resilience Alliance in 2019. We used the variables to
annotate qualitative aspects such as use of the adaptive cycle, identification
of its phases, whether the paper is conceptual, modeling or an empirical study.
When empirical, we recorded the temporal and spatial scales of the case study.
We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used.

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 4in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25. While 50 topics is marginally better, for visualization purposes we restrict our analysis to 25. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}
```
Comparing algorithm performance revealed that Gibbs sampling was the best fit
for the data (Fig \ref{fig:algorithm}). As a rule of thumb, an ideal method
should maximize entropy and the log-likelihood estimation while minimizing
perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and likelihood
with our data when compared to other alternatives. The second best was the
variational-expectation maximisation algorithm (VEM) when $\alpha$ was not set
constant. $\alpha$ is a hyper parameter that weights the evenness of topic
distribution. A lower $\alpha$ than default values indicates that the documents
can be described by rather fewer topics, or that its distribution is very
uneven. In fact, we observed that increasing the number of topics from 5 to 100
topics did increase entropy at expense of reducing $\alpha$, meaning that
despite the larger number of topics, the main content of a document is still
captured by a few of them. The log-likelihood maximization stopped at around
25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our model fit. See FigS1 for a panel of time series per topics}
\label{fig:topics25}
\end{figure*}
```
A topic is a set of words that are ranked according to a probability that they
represent an underlying content of the document (Fig \ref{fig:topics25}). For
example, the words "resilience", "vulnerability", "risk" and "disaster" have a
high probability to capture the content of topic 25. Papers early in our time
series (2001-3) have high content largely dominated by topics 25 on resilience,
24 on the adaptive cycle, and 18 on adaptive governance and management. Towards
the end of the time series, topic 3 on urban systems and 9 on local communities
and knowledge have spikes up to 9% of the content of each year (2019-2020). For
comparison, if all topics were equally represented in the content, they would
have 4% in the corpus (grey line in Fig \ref{fig:topics25}B).

Despite fluctuations, most topics showed a relatively constant level of interest
over time (Fig \ref{fig:topics25}, Fig \ref{fig:figS1}). We did not observe
strong trends, but some topics have gained a small amount of attention. For
example, topic 1 on sustainability, and topic 10 on ecosystem services appeared
consistently across time. In contrast, research on business innovation (topic
20), urban infrastructure (topic 3), and archaeology (topic 5) have gained
attention in recent years. Topic 8 was an outlier, with a selection of papers
ranking high on content related to network infrastructure and performance,
possibly from engineering disciplines. It was the only topic with a set of
papers that was clearly distinct from the rest of the collection, and showed a
decreasing trend over time.

The human-coded document analysis revealed that the most common feature of
panarchy in the literature was the adaptive cycle (81%, N = 42) followed by
cross-scale interactions (Fig \ref{fig:qual}). Poverty and rigidity traps were
less studied features in our sample (14% and 19% respectively), even when
accounting for slightly different terminology such as "lock-ins". The bulk of
the papers analyzed were conceptual papers, many without defined methods or discernible
research question. Roughly half of the papers were conceptual or only used
panarchy as a metaphor. Over 40% of papers in our sample went a step further and
used panarchy as an empirical construct, for example by attempting to identify
the phases of the adaptive cycle (61%). Out of the 22 empirical cases, 6 were at
the time scale of centuries, 7 on decades, 6 on years, and 1 in weeks.
Spatially, 4 were at a city scale, 17 regional, and 1 national. Empirical papers
predominantly used qualitative methods (77%, N = 22) and were generally
retrospective historical reviews.

Several of the papers reviewed attempted to identify the adaptive cycle either
through qualitative or quantitative methods. Identification of adaptive cycles
has found applications in a wide range of disciplines and research problems,
from delimitation of periods in archaeology and anthropology [@Redman:2003bd],
to financial crises in Europe [@Castell:2020dg], to traffic jams in China
[@Zeng:2020hu]. Some recent empirical tests of the adaptive cycle used innovative
types of data and methods. For example, information transfer methods based
on entropy have been suggested to approximate relevant components of a system
and the empirical proxies of the axes within which the adaptive cycle unfolds:
potential, connectedness, and resilience [@Castell:2020dg]. Percolation methods
combined with big data have been proven useful for testing hypotheses of regime
shifts in urban systems (originally proposed in Panarchy [@Gunderson:2002vk]),
and deriving the temporal and spatial scales at which the adaptive cycle emerges
[@Zeng:2020hu].

@@

The high level of conceptualization but lack of empirically grounded testing of panarchy theory
came across as one of the major limitations. Here we do not refer to positivism as the only plausible way of theory development, but acknowledge that other equally valid
approaches exist. From the philosophical point of view of what makes a
theory a scientific theory, the plurality of approaches should have mechanisms
in place to avoid confirmation bias, allow the emergence of alternative
explanations, and identify the circumstances under which a theory may fail
[@Hansson:2017tr]. Many of the papers we read confirmed but not challenge the
expectations that Panarchy put forward, though a few of the most recent works did
falsify some of the ideas proposed --e.g. the back loop is faster that the front
one [@Castell:2020dg]. We believe more work is needed to better understand the circumstances where
Panarchy works, and when it falls short. In other words, to refine the scientific
framework into a scientific theory.

@@

The criticism of over-conceptualization is not unique to the literature that engage with
Panarchy ideas. Here by over-conceptualization we refer to the problem of
scientific jargon, where different communities use slightly different wordings
to refer to the same phenomenon; reducing the chances that lessons from a
disciplines percolates to another. An example from Panarchy is the endorsement
of ideas such as creative destruction (from economics) or the theory of
intermediate disturbances (from ecology). The latter has been debunked in the
ecological literature but might still been used within Panarchy thinking
[@Fox:2013br]. A recent review of sustainability science mapped the different
schools of thought that the discipline has developed over the last decades
[@Clark:2020jd] and reached a similar conclusion: too many conceptual
frameworks have been developed, but there have been too few empirical attempts to test the
frameworks against data and falsify hypotheses. The review also emphasizes the problem
of measurement and observation [@Clark:2020jd]. In the context of panarchy,
recent work has developed an information theory based approach that enables the
identification of adaptive cycles [@Castell:2020dg]. The authors identify phases
of the adaptive cycle in the European financial crises and in grassland
ecosystems, but fail to find support for the difference in speed between the
forward and backward loops originally proposed in Panarchy.

>>>>

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry Peterson^1^
    | \small
    | \footnotesize ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \footnotesize ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \footnotesize ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \footnotesize ^4^Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \footnotesize ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \footnotesize ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, 
    | \footnotesize Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \footnotesize ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA
abstract: |
  How do social-ecological systems change over time? In 2002 Holling and colleagues proposed the concept of Panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines, and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by combining qualitative methods and machine learning. Document analysis was used to code panarchy features that are commonly used in the scientific literature (N = 42), a qualitative analysis that was complemented with topic modeling of 2177 documents. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.
bibliography: [panarchy.bib, packages.bib]
csl: "/Users/juanrocha/Documents/styles/ecology.csl"
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
---

```{r message = FALSE, warnings=FALSE, include = FALSE}
library(here)
library(tidytext)
library(tidyverse)
library(tm)
library (topicmodels)
library(lda)

load(here("data/dtm.Rdata"))
load(here("data/qual_coding.Rdata"))
```

@@

Panarchy proposes that it is useful to conceptualize systems in terms of
interacting adaptive cycles. The adaptive cycle was an idea first proposed by
Holling based upon his experience working and studying managed ecosystems
[@biosphere:hJ4c_Ua8]. It was meant to be a conceptual tool that focused
attention on processes of destruction and reorganization, which had been
neglected in comparison to those of growth and conservation.

An adaptive cycle alternates between long periods of system aggregation,
connection, and accumulation, and shorter periods of disruption and
reorganization. The adaptive cycle exhibits four phases arranged into two major
loops (Fig. \ref{fig:fig1}). The phases are exploitation ($r$), conservation
($K$), release ($\Omega$), and reorganization ($\alpha$), the latter
characterized by events of creative destruction (Fig. \ref{fig:fig1}). The first
loop, often referred to as the front-loop, from $r$ to $K$, is the slow,
incremental phase of growth and accumulation. The second loop, referred to as
the back-loop, from $\Omega$ to $\alpha$, is the rapid phase of reorganization
leading to renewal. A system going into the back-loop can either remain in a
similar form, or may transition to a new type of system with new boundaries and
key components [@Holling:2002wu]. Resilience researchers proposed that tensions
between demands for organization and efficiency, versus demands for novelty and
diversity, drive adaptive cycle dynamics in many different types of complex
systems [@Holling:2002wu].

Panarchy is not a theory of what it is, but a metaphor of what might be
[@Gunderson:2002vk]. It is not a predictive tool, but aims to understand
adaptive change. The adaptive cycle is hypothesized to exist within a
three-dimensional space defined by three properties: potential, connectedness,
and resilience [@Gunderson:2002vk]. Potential refers to the capital available to
the system, for example nutrients and carbon captured by a forest, or human
capital - skills and knowledge accumulated to run the economy. Connectedness is
a proxy of the structure of the system; it is the network of interactions and
strength between its elements. Resilience is the capacity of any system to
absorb disturbance and reorganize while undergoing change so as to retain
essentially the same identity, as determined by its function, structure, and
feedbacks [@Anonymous:2016fv].

@@

The three-dimensional space has corners with attractors that can cause the
adaptive cycle to stagnate: poverty and rigidity traps [@Gunderson:2002vk].
Poverty traps are described in panarchy as maladaptive states where potential,
connectedness and resilience are low. Poverty traps are a series of feedback
mechanisms that reinforce impoverished states [@Bowles:2006hx; @Maru:2012dm;
@Allison:2004te], limiting the system's capacity to innovate and increase
potential. The opposite corner, where potential, connectivity and resilience are
high, is another maladaptive space called a rigidity trap. In that corner there
is little space for experimentation and innovation. Examples include systems
where ecological resilience has been extensively replaced by artificial
processes to maintain the system such as levees, flood barriers, or chemical
control of pests [@Holling:1996wx].

Panarchy offers a rich conceptual framework for understanding environmental
problems. While inspired by several mathematical constructs (e.g. cycles, traps,
scaling laws), it is general enough to invite scholars from multiple
disciplinary backgrounds, ontologies and epistemologies, to collaborate around
research questions and applied problems. As such, it is useful as a boundary
object and can be used empirically or metaphorically. The book itself presents a
series of case studies where geographers, economists, political scientists, and
ecologists have demonstrated the utility of the framework to their area of
research.

However, the extent to which these concepts have gained empirical support beyond
the metaphor, or how they have evolved over time, remain open questions. For
example, the adaptive cycle is a useful metaphor to look back at history and
organize events and periods around phases, but it remains challenging to
identify the phase of a system in real time, or to use panarchy to project
future trajectories. Scientific theories are bodies of knowledge that answer the
question why? The answer implies causal reasoning. Regardless of discipline, a
theory should therefore provide a series of testable expectations that help
distinguishing the situations when it is a good explanation of reality, versus
situations when it falls short. The practice of science implies avoiding
confirmation bias and testing when explanations are valid representations of
reality. For panarchy to become a useful theory in sustainability science, it
needs to stand the rigor of testing, and being able to explain phenomena in
social-ecological systems across different contexts. We identify progress and
knowledge gaps in the scientific practice inspired by panarchy.

@@

To answer these questions we combined an automated literature review based on
topic modeling [@Blei:2012dk; @Griffiths:2004ey] with human-coded document
analysis [@Bryman:2008va]. We first describe the data curation and how it was
used in the topic models. Then we introduce the qualitative approach.

*Data*: We used the Web of Science, Scopus, and GoogleScholar to survey academic
literature that has used or referenced works that trace back to Panarchy
[@Gunderson:2002vk]. We extracted complete records from the Scopus database that
matched the search for *"panarchy OR adaptive cycle"* (N = 595), or the search
*"(panarchy OR adaptive cycle) AND resilience"* (N = 278). The data was combined
with all papers (N = 1923) that cited the inaugural paper that introduced the
book to the academic community [@Holling:2001cu]. Records with missing abstracts
were dropped (N = 191), and records with missing year were set to 2020 given
that they are accepted manuscripts with digital object identifier (doi)
scheduled to be published later in 2020 or 2021. Our final sample is
`r dim(dtm)[1]` documents.

To prepare the data for topic modeling, we constructed a document term matrix
with documents in rows (N=`r dim(dtm)[1]`) and words in columns
(N=`r dim(dtm)[2]`). Here our unit of analysis for the document are the
abstracts retrieved, and the matrix contains the count of the number of
occurrences of each word per abstract. We removed stop words (e.g. "the", "a")
and digits from the matrix, as well as a list of words that were over
represented in our data and are common in the scientific literature but are
unrelated to the papers' topics (e.g. "paper", "study", "aim"). The lexicon of
stop words has 1149 entries, curated by various research groups, and accessible
in the `R` language under the packages `tm` and `tidytext` [@R-topicmodels;
@R-tm; @R-NLP; @tm2008; @R-tidytext], for example by `tidytext::stop_words`. The
data was also lemmatized using the hash lemmas dataset available at
[\<http://www.lexiconista.com\>](http://www.lexiconista.com/). Lemmatization is
the process of simplifying words to their root lemmas (e.g. *teaching* and
*taught* are lemmatized to *teach*).

@@

The underlying statistical technique for this machine learning approach is
called Latent Dirichlet Allocation (LDA) [@Blei:2003tn]. It allocates
probabilities to latent variables (topics) based on the distribution of words in
text data, assuming a multivariate continuous (Dirichlet) distribution. We
compared three LDA algorithms: correlated topic models (CTM), variational
expectation maximization (VEM), and Gibbs sampling (Gibbs), by assessing their
performance against their log-likelihood estimation, entropy and perplexity
[@Grun:2011tb]. Entropy is a measure of order or disorder of a system; in the
context of topic models it measures how evenly the topic distribution is spread.
Perplexity measures the uncertainty of predicting a single word, so if the model
performance were the same as random, perplexity would approximate the vocabulary
size (N = `r dim(dtm)[2]` words). These performance metrics enable us to choose
which algorithm best fits our data, what is the optimal number of topics to fit,
and how to avoid overfitting.

*Document analysis*: We complemented our unsupervised approach with document
analysis [@Bryman:2008va] by coding an additional set of categorical variables
for a sample of papers (N = `r nrow(qual)`). The papers read are part of the
corpus used for topic modelling. We read papers avoiding biasing our selection
to high cited papers, specific journals, older papers or particular disciplines
or methods.

Based on the reading of Panarchy, we first developed a series of categorical
variables that captured the main components of the scientific framework. The
variables were presented and received feedback at the science meeting of the
Resilience Alliance in 2019. We used the variables to annotate qualitative
aspects such as use of the adaptive cycle, identification of its phases, whether
the paper is conceptual, modeling or an empirical study. When empirical, we
recorded the temporal and spatial scales of the case study. We also
distinguished empirical cases from empirical constructs. An empirical case is
when a study is trying to test an idea or proposition put forward by Panarchy
(e.g. a cross-scale interaction, or that the back loop is faster than the forth
loop). An empirical construct is when Panarchy is used as inspiration to
interpret a case study, typically with a historical component to it (e.g.
identify the phases of adaptive cycle); but it does not find support for or
against a Panarchy preposition. That is, there was no mechanism in place to
avoid confirmation bias.

We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used. We did a preliminary scoping and coding of papers, and
through a iterative process we kept track of potential fields under reviewed by
our corpus, or simply absent of it. We coded papers until we reached conceptual
saturation, that is, additional papers were not proving additional categories of
the already coded. The code book is available at
[doi.org/10.6084/m9.figshare.13490919](http://doi.org/10.6084/m9.figshare.13490919).

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 4in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25 and 100. While 50 topics is marginally better. We choose 25 because the $\alpha$ parameter suggest preferring lower numbers, or a more parsimonious solution to avoid over fitting. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}
```
Comparing algorithm performance revealed that Gibbs sampling was the best fit
for the data (Fig \ref{fig:algorithm}). As a rule of thumb, an ideal method
should maximize entropy and the log-likelihood estimation while minimizing
perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and likelihood
with our data when compared to other alternatives. The second best was the
variational-expectation maximisation algorithm (VEM) when $\alpha$ was not set
constant. $\alpha$ is a hyper parameter that weights the evenness of topic
distribution. A lower $\alpha$ than default values indicates that the documents
can be described by rather fewer topics, or that its distribution is very
uneven. In fact, we observed that increasing the number of topics from 5 to 100
topics did increase entropy at expense of reducing $\alpha$, meaning that
despite the larger number of topics, the main content of a document is still
captured by a few of them. The log-likelihood maximization stopped at around
25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). In (B), a reference dotted horizontal line marks the level where all topics are equally dominant. Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our model fit. Fig \ref{fig:figS1} expands (B) presenting time series for each topic. All time series exclude 2021 to account for the artificial decline of papers at the end of the time series as shown in A.}
\label{fig:topics25}
\end{figure*}
```
A topic is a set of words that are ranked according to a probability that they
represent an underlying content of the document (Fig \ref{fig:topics25}). For
example, the words "resilience", "adaptive", or "capacity" have a high
probability to capture the content of topic 14. Papers early in our time series
(2001-3) have high content largely dominated by topics 14 on resilience, 9 on
resource management, and 23 on cross-scale interactions. Towards the end of the
time series, topic 7 on urban systems and 25 on local communities and knowledge
have spikes up to 9% of the content of each year (2019-2020). For comparison, if
all topics were equally represented in the content, they would have 4% in the
corpus (grey line in Fig \ref{fig:topics25}B). The topics presented lack names
for reproducibility purposes. While here we present e.g. topic 7 as the urban
topic, if a researcher replicates the analysis that person will find the same
list of words and probability distributions listed under a topic with a
different number. That is because the labeling of topics is random and we do not
want to bias the reader to over interpret the word patterns found.

```{=tex}
\begin{figure*}[h]
\centering
\includegraphics[width = 6in, height = 5in]{SM_topics25_timeline}
\caption{\textbf{Proportion of papers per year per topic.} Each line represent the average proportion of content for a given year over all papers published in the same year. Fig \ref{fig:figS2} complements with the proportion of topics for each paper in our sample.}
\label{fig:figS1}
\end{figure*}
```
Despite fluctuations, most topics showed a relatively constant level of interest
over time (Fig \ref{fig:topics25}, \ref{fig:figS1}). We did not observe strong
trends, but some topics have gained a small amount of attention. For example,
topic 11 on sustainability, and topic 15 on ecosystem services appeared
consistently across time. In contrast, research on innovation (topic 18), urban
infrastructure (topic 7), and local communities (topic 13) have gained attention
in recent years. Topic 6 was an outlier, with a selection of papers ranking high
on content related to network infrastructure and performance, possibly from
engineering disciplines. It was the only topic with a set of papers that was
clearly distinct from the rest of the collection, and showed a decreasing trend
over time.

The human-coded document analysis revealed that the most common feature of
panarchy in the literature was the adaptive cycle (81%, N = `r nrow(qual)`)
followed by cross-scale interactions (Fig \ref{fig:qual}). Poverty and rigidity
traps were less studied features in our sample (14% and 19% respectively), even
when accounting for slightly different terminology such as "lock-ins". 52.4% of
the papers analyzed were conceptual papers, many without defined methods or
discernible research question. Roughly half (47%) of the papers were conceptual
or only used panarchy as a metaphor. Over 40% of papers in our sample went a
step further and used panarchy as an empirical construct, for example by
attempting to identify the phases of the adaptive cycle (61%). Out of the 22
empirical cases, 6 were at the time scale of centuries, 7 on decades, 6 on
years, and 1 in weeks. Spatially, 4 were at a city scale, 17 regional, and 1
national. Empirical papers predominantly used qualitative methods (77%, N = 22)
and were generally retrospective historical reviews.

Several of the papers reviewed (81%) attempted to identify the adaptive cycle
either through qualitative or quantitative methods. Identification of adaptive
cycles has found applications in a wide range of disciplines and research
problems, from delimitation of periods in archaeology and anthropology
[@Redman:2003bd], to financial crises in Europe [@Castell:2020dg], to traffic
jams in China [@Zeng:2020hu]. Some recent empirical tests of the adaptive cycle
used innovative types of data and methods. For example, information transfer
methods based on entropy have been suggested to approximate relevant components
of a system and the empirical proxies of the axes within which the adaptive
cycle unfolds: potential, connectedness, and resilience [@Castell:2020dg].
Percolation methods combined with big data have been proven useful for testing
hypotheses of regime shifts in urban systems (originally proposed in Panarchy
[@Gunderson:2002vk]), and deriving the temporal and spatial scales at which the
adaptive cycle emerges [@Zeng:2020hu].

@@

@@

The high level of conceptualization but lack of empirically grounded testing of
panarchy theory came across as one of the major limitations, given our sample of
papers reviewed (N=`r nrow(qual)`). What makes a theory a scientific theory is
that the plurality of scientific approaches (e.g. positivism, constructivism)
should have mechanisms in place to avoid confirmation bias, allowing the
emergence of alternative explanations, and identify the circumstances under
which a theory may fail [@Hansson:2017tr]. Many of the papers we read confirmed
but did not challenge the expectations that Panarchy put forward, though a few
of the most recent works did falsify some of the ideas proposed --e.g. the back
loop is faster that the front one [@Castell:2020dg]. We believe more work is
needed to better understand the circumstances where Panarchy works, and when it
falls short. In other words, to refine the scientific framework into a
scientific theory.

The criticism of over-conceptualization is not unique to the literature that
engages with Panarchy ideas. Here by over-conceptualization we refer to the
problem of scientific jargon, where different communities use slightly different
wordings to refer to the same phenomenon; reducing the chances that lessons from
a discipline percolates to another. An example from Panarchy is the endorsement
of ideas such as creative destruction (from economics) or the theory of
intermediate disturbances (from ecology). The latter has been debunked in the
ecological literature but might still been used within Panarchy thinking
[@Fox:2013br]. A recent review of sustainability science mapped the different
schools of thought that the discipline has developed over the last decades
[@Clark:2020jd] and reached a similar conclusion: too many conceptual frameworks
have been developed, but there have been too few empirical attempts to test the
frameworks against data and falsify hypotheses. The review also emphasizes the
problem of measurement and observation [@Clark:2020jd]. In the context of
panarchy, recent work has developed an information theory based approach that
enables the identification of adaptive cycles [@Castell:2020dg]. The authors
identify phases of the adaptive cycle in the European financial crises and in
grassland ecosystems, but fail to find support for the difference in speed
between the forward and backward loops originally proposed in Panarchy.
